---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
As a first step, read.csv() function is used to load the activity.csv file to a variable called ractivity. Take a look at the data using head() and tail() functions.
```{r}
ractivity <- read.csv("activity.csv")
head(ractivity)
tail(ractivity)
```
Use summary() and str() functions to view more information about the data.
```{r}
summary(ractivity)
str(ractivity)
```


## What is mean total number of steps taken per day?
Assign to a new variable without the missing values (NAs) and use tapply() to get total steps per day. 
```{r}
rmna_activity <- ractivity[!is.na(ractivity$steps),]
tapply(rmna_activity$steps,rmna_activity$date, sum)
```
Let's make a histogram of the total number of steps per day.
```{r}
 hist(tapply(rmna_activity$steps,rmna_activity$date, sum),main ="Histogram of Total number of steps taken", xlab="Number of Steps",col=c(3:7))
```

Apply mean() and median() functions on the total steps per day.
```{r}
mean(tapply(rmna_activity$steps,rmna_activity$date, sum),na.rm = TRUE)
median(tapply(rmna_activity$steps,rmna_activity$date, sum),na.rm = TRUE)
```


## What is the average daily activity pattern?
Using aggregate() function to get average steps taken across all days and make a time series plot. Convert the interval into hours to get a proportionate scale.

```{r}
avstep <- aggregate(steps ~ interval, mean, data=ractivity, rm.na =TRUE)
avstep$hr <- (avstep$interval %/% 100) + (avstep$interval %% 100)/60
plot(avstep[,3],avstep[,2], type="l", xlab = "Interval (hours)", ylab="Average Steps", main ="Average Steps Taken Across All Days", col=3)
```

To find the 5-minute interval, on average across all the days in the dataset,which contains the maximum number of steps,
```{r}
avstep[round(avstep[,2],2) == round(max(avstep[,2]),2),]
```
## Imputing missing values

To find the missing values (NAs) in the data set,
```{r}
sum(is.na(ractivity[,1]))
```
To assign average of average number of steps taken for 5-minute interval across all days to the missing values,
```{r}
nactivity <- ractivity
nactivity[is.na(nactivity[,1]),1] <- mean(avstep[,2])
```

Plot a histogram with newly created data set
```{r}
 hist(tapply(nactivity$steps,nactivity$date, sum),main ="Histogram of Total number of steps taken", xlab="Number of Steps",col=c(3:7))
```

Apply mean() and median() functions on the total steps per day on new data set
```{r}
mean(tapply(nactivity$steps,nactivity$date, sum))
median(tapply(nactivity$steps,nactivity$date, sum))
```
Since the average is assigned to all the missing value, the new mean is the same as the mean from old data set.


## Are there differences in activity patterns between weekdays and weekends?

Add a new column to the data set nactivity and assign "Weekday" and "Weekend" accordingly. Convert that column as factor. Generate average steps taken across all days using aggregate function and assign to a variabl called navstep. Convert interval into hours and add as a new column to navstep.  

```{r}
nactivity$Wday <- rep(NA, nrow(nactivity))
nactivity[weekdays(as.Date(nactivity[,2])) %in% c("Saturday","Sunday"),4] <- "Weekend"
nactivity[!weekdays(as.Date(nactivity[,2])) %in% c("Saturday","Sunday"),4] <- "Weekday"
nactivity$Wday <- as.factor(nactivity$Wday)
navstep <- aggregate(steps ~ interval+Wday, mean, data=nactivity, rm.na =TRUE)
navstep$hr <- (navstep$interval %/% 100) + (navstep$interval %% 100)/60
```

Then make a panel plot using Lattice plotting system.
```{r}
library(lattice)
xyplot(steps ~ hr | Wday, data = navstep, layout = c(1,2), type = "l", xlab = "Interval (hours)",ylab = "Number of Steps")
```

We can see from the graph that during the weekend the activity is spread out the whole day.  During the weekday, the activity level peaks just before the office hour --most probably caused by the commute to work.
